The Weakly Nonlinear Magnetorotational Instability in a Taylor Couette Flow

### 2016-07-14 paper/widegap_mri_paper.tex
some hmri intro to widegap paper added

@@

@@

In this work we will focus our findings on two fiducial parameter sets, one for the standard MRI (hereafter SMRI) where $\xi = 0$ and one for the helical MRI (hereafter HMRI). Our fiducial parameters are described in Table \ref{table:parameters}.

\begin{table}[ht]\label{table:parameters}
\normalsize
\caption{Fiducial parameters for MRI runs} 
\centering
%\resizebox{\textwidth}{!}
{\begin{tabular}{rlrrrrrrr}
  \hline
 & $\xi$ & $Pm$ & $\Omega_2/\Omega_1$ & $\beta$ & X \\ 
  \hline\hline
Standard MRI & 0 & 25 & 0 & 0 & 0 \\ 
Helical MRI & 4 & 0 & 0\\ 
   \hline
\end{tabular}}
\end{table}

@@

where J is the Jacobian $J(x, y) \equiv \partial_z x \partial_r y - \partial_r x \partial_z y$. Note that in the above, $\nabla^2 x \equiv \partial_r^2 x + \partial_z^2 x + \frac{1}{r} \partial_r x$. Equations \ref{eq:Psi_perturbed} - \ref{eq:Bphi_perturbed} are nondimensionalized, where lengths have been scaled by $r_0$, velocities by $r_0 \Omega_0$, densities by $\rho_0$, and magnetic fields by $B_0$; where $B_0$ appears in the above it is formally unity. $\Omega_0 = \Omega(r_0)$ is the rotation rate at the center of the channel. We introduce the Reynolds number $\reye = \Omega_0 r_0^2/\nu$, the magnetic Reynolds number $\reym = \Omega_0 r_0^2 / \eta$, and a plasma beta parameter $\beta = \Omega_0^2 r_0^2 \rho_0/B_0^2$. If we define the dimensional cylindrical coordinate $r = r_0(1 + \delta x)$, the narrow gap equations are recovered in the limit $\delta \rightarrow 0$.

@@

@@

>>>>

Despite the challenges inherent in studying the MRI, much progress has been made nu

@@

Because we include curvature terms, our theory is also relevant to the helical magnetorotational instability (HMRI). Discovered by \citeauthor{Hollerbach:2005tr} (\citeyear{Hollerbach:2005tr}), the HMRI is an overstability in which the background magnetic field is helical. The HMRI currently occupies a special place in the MRI puzzle. The HMRI has been proposed as a method of awakening angular momentum transport in the ``dead zones" of protoplanetary disks where the magnetic Prandtl number ($Pm = \nu/\eta$) becomes very small. However the rotation profiles needed to excite HMRI may be prohibitively steeper than Keplerian depending on the boundary conditions, and so its role in astrophysical disks is currently debated (\citei{Liu:2006}, \citei{Rudiger:2007}, \citei{Kirillov:2013}). However, the HMRI is significantly easier to excite in a laboratory setting, and has already been detected in the laboratory by the Potsdam Rossendorf Magnetic Instability Experiment (PROMISE; \citei{Stefani:2006iv}, \citei{Stefani:2009hp}).

We explore the weakly nonlinear HMRI

@@

In this work we will focus our findings on two fiducial parameter sets, one for the standard MRI (hereafter SMRI) where $\xi = 0$ and one for the helical MRI (hereafter HMRI). We choose the SMRI parameters to be comparable to the case explored in \citei{Umurhan:2007hs}, with the geometric parameters of \citei{Goodman:2002ix}. The HMRI parameters were chosen to be comparable to \citei{Hollerbach:2005tr}. Our fiducial parameters are described in Table \ref{table:parameters}.

\begin{table}[ht]
\normalsize
\caption{Fiducial parameters for MRI runs} \label{table:parameters}
\centering
%\resizebox{\textwidth}{!}
{\begin{tabular}{rlrrrrrrr}
  \hline
 & $\xi$ & $Pm$ & $\beta$ & $\Omega_2/\Omega_1$ & $R_1/R_2$ \\ 
  \hline\hline
Standard MRI & 0 & 1E-3 & 25 & 0.18 & 0.33 \\ 
Helical MRI & 4 & 1E-6 & 1.7E-2 & 0.27 & 0.5\\ 
   \hline
\end{tabular}}
\end{table}

@@

where J is the Jacobian $J(f, g) \equiv \partial_z f \partial_r g - \partial_r f \partial_z g$. Note that in the above, $\nabla^2 f \equiv \partial_r^2 f + \partial_z^2 f + \frac{1}{r} \partial_r f$. Equations \ref{eq:Psi_perturbed} - \ref{eq:Bphi_perturbed} are nondimensionalized, where lengths have been scaled by $r_0$, velocities by $r_0 \Omega_0$, densities by $\rho_0$, and magnetic fields by $B_0$; where $B_0$ appears in the above it is formally unity. $\Omega_0 = \Omega(r_0)$ is the rotation rate at the center of the channel. We introduce the Reynolds number $\reye = \Omega_0 r_0^2/\nu$, the magnetic Reynolds number $\reym = \Omega_0 r_0^2 / \eta$, and a plasma beta parameter $\beta = \Omega_0^2 r_0^2 \rho_0/B_0^2$. If we define the dimensional cylindrical coordinate $r = r_0(1 + \delta x)$, the narrow gap equations are recovered in the limit $\delta \rightarrow 0$.

@@

\bibitem[{Goodman \& Ji(2002)}]{Goodman:2002ix}
Goodman, J. \& Ji, H., 2002, JFM, 462, 365

\bibitem[{Hollerbach \& R\"udiger(2005)}]{Hollerbach:2005tr}
Hollerbach, R. \& R\"udiger, G., 2005, Phys. Rev. Lett. 95, 124501

\bibitem[{Kirillov \& Stefani(2013)}]{Kirillov:2013}
Kirillov, O, \& Stefani, F, 2013, Phys. Rev. Lett. 111, 061103

@@

\bibitem[{Liu et al.(2006)}]{Liu:2006}
Liu, W, Goodman, J, Herron, I, and Ji, H, 2006, Phys. Rev. E 74, 056302

\bibitem[{R\"udiger \& Hollerbach(2007)}]{Rudiger:2007}
R\"udiger, G, \& Hollerbach, R, Phys. Rev. E 76, 068301

\bibitem[{Stefani et al.(2006)}]{Stefani:2006iv}
Stefani, F, Gundrum, T, Gerbeth, G, RŸdiger, G, Schultz, M, Szklarski, J, Hollerbach, R, Phys Rev Lett, 2006, 97, 184502

\bibitem[{Stefani et al.(2009)}]{Stefani:2009hp}
Stefani, F, Gerbeth, G, Gundrum, T, Hollerbach, R, Priede, J, R\"udiger, G, and Szklarski, J, Phys Rev E, 80, 066303
